White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2017-09-09 papers/wdmerger_II/paper.tex
Add note on threshold for timestep limiter for X
Clarify the understanding of WD collision rates

Binary white dwarf (WD) systems have long been thought to be potential progenitors
of Type Ia supernovae (SNe Ia) \citep{ibentutukov:1984,webbink:1984}. Recently there
has been a growing amount of observational evidence that double WD systems play
an important role in explaining at least some SNe Ia \citep{maoz:2014}. These systems
may be isolated or they may be in hierarchical triple systems, where a WD binary
system in a tight inner orbit is gravitationally coupled to a third, outer star.
In either case the binary WD pair can merge due to the emissions of gravitational waves,
but in the hierarchical triple case, dynamical interactions between the WD binary and
the outer star can prompt mergers as well \citep{thompson:2011,hamers:2013}. The mergers
occur due to perturbations in the eccentricity of the binary orbit caused by the outer
star, which significantly increase the gravitational wave emission of the system. In a
subset of these mergers the eccentricity may be driven to such a high value that the
merger resembles a head-on collision of the binary WDs. Whether the rate of these collisions
may be high enough to significantly contribute to the observed SN Ia rate is uncertain,
though \cite{katzdong:2012} suggest that this is plausible.

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_e$ parameter in the
literature is in the range of 0.2 or 0.3, while to our knowledge a limiter based on
$f_X$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed in 
\autoref{sec:parameters:timestepping}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

>>>>

Binary white dwarf (WD) systems have long been thought to be potential progenitors
of Type Ia supernovae (SNe Ia) \citep{ibentutukov:1984,webbink:1984}. Recently there
has been a growing amount of observational evidence that double WD systems play
an important role in explaining at least some SNe Ia \citep{maoz:2014}. These systems
may be isolated or they may be in hierarchical triple systems, where a WD binary
system in a tight inner orbit is gravitationally coupled to a third, outer star.
In either case the binary WD pair can merge due to the emissions of gravitational waves,
but in the hierarchical triple case, dynamical interactions between the WD binary and
the outer star can prompt mergers as well \citep{thompson:2011,hamers:2013}. The mergers
occur due to perturbations in the eccentricity of the binary orbit caused by the outer
star, which significantly increase the gravitational wave emission of the system. In a
subset of these mergers the eccentricity may be driven to such a high value that the
merger resembles a head-on collision of the binary WDs. Whether the rate of these collisions
may be high enough to significantly contribute to the observed SN Ia rate is debated:
\cite{katzdong:2012} suggest that this is plausible, but \cite{hamers:2013} and
\cite{papish:2015} suggest that the rate is relatively low. Even if the rate is low,
if at least some SNe Ia are caused by WD collisions then they are worth understanding.

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{X}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_e$ parameter in the
literature is in the range of 0.2 or 0.3, while to our knowledge a limiter based on
$f_X$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed in 
\autoref{sec:parameters:timestepping}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.
